\bf Adaption of the Chumbley Score to matching of bullet striation marks

### 2018-04-18 doc/paper.Rmd
expected failure rate

## Motivation
Same source analyses are a major part of an Forensic Toolmark Examiner's job. In current practice examiners  make these comparisons by visual inspection under a comparison microscope and come to one of the following four conclusions: identification, inconclusive, elimination or unsuitable for examination~\citep{afte-toolmarks1998}. These conclusions are made on the basis of "unique surface contours" of the two toolmarks being in "sufficient agreement" \citep{afte-toolmarks1998}. AFTE describes the term "sufficient agreement" as the possibility of another tool producing the markings under comparison, as practically impossible \citep{afte-toolmarks1998}. This subjectivity in the assessment as well as the lack of error rates are the main points of criticisms  first raised by the National Research Council in 2009 \citep{NAS:2009} and later emphasized further by the President's Council of Advisors on Science and Technology \citep{pcast2016}.

@@

Same-shift pairs of length $w_v$ are defined in \citet{hadler} as all 
pairs $(s_i^{(1)}, s_i^{(2)})$ for all integer values $i$ for which both $s_i^{(1)}$ and  $s_i^{(2)}$ are defined. Similarly, different-shift pairs are defined as $(s_i^{(1)}, s_{-i}^{(2)})$ for all $i$ where both $s_i^{(1)}$ and  $s_{-i}^{(2)}$ are defined.

@@

#### endedit

\hh{end of intro: remaining paper is structured as follows: introduce to the data we get from confocal microscopy, introduce to profiles and signatures.
Introduce to chumbley score method, apply chumbley and discuss results ....
}

- scans available: NIST database (citation), Hamby 44 and Hamby 252 (Hamby citation)
- move figure up
- discuss cross section, profiles and signatures
-->

@@

<!-- As mentioned earlier, subject bias and error rate determination have been a long standing issue in firearm examination \citep{NAS:2009}. This issue of removing subjectivity and making firearm analysis objective  was also raised in the \citet{pcast2016} report. For an objective analysis, scientific principle testability and error rate determination are considered to be defining aspects.  Identification of the uncertainities associated with a method of comparison and coming up with a systemized method of quantification, is therefore important. Associating mathematical probabilites in a systematic manner to these methods are one such way of quantification.  -->
<!-- In a previous study conducted by \citet{aoas} a machine learning based algorithm was developed for same source matching of bullets and error rates were discussed using the database from the Hamby Study \citep{hamby}. The method proposed by \citet{chumbley} (and later improved by \citet{hadler}) also provides a means to determine error rates and claims to reduce subject bias. Therefore, giving us a strong motivation to explore the adaptibility of the Chumbley score methodology to bullets. In this paper, we take this toolmark based methodology and try to apply and adapt it to bullets. After this, we consequently discuss about the efforts in doing so, along with the associated error rates. The data used in this paper also belongs to the Hamby Study \citep{hamby}. This gives us a common platform for comparing the performance of the chumbley method on bullets with the already existing method proposed by \citet{aoas} for bullets. As a brief overview, \citet{chumbley}, in their paper, compare two toolmarks with the intention of determining if it comes from the same tool.  They use an empirical based setup to validate their proposed algorithm and quantitative method which calculates a U-statisic for the purpose of classification of toolmarks as matching or non-matching. The data for their study was obtained from 50 sequentially manufactured screwdriver tips, and preselected comparison window sizes were given as inputs to the algorithm. The algorithm then compares the two toolmarks and comes up with a Mann-Whitney U-statistic and an associated p-value to designate them as matches or non-matches. The performance for every 100 comparisons, of the algorithm proposed by \citet{chumbley} and the improvement proposed by \citet{hadler} are listed in the table below. -->

<!-- ```{r, results='asis', include=FALSE} -->
<!-- library(knitr) -->
<!-- library(kableExtra) -->
<!-- library(xtable) -->

<!-- a1<- matrix(NA, ncol = 3, nrow = 2) -->
<!-- toolmark1.confusion<- data.frame(a1) -->
<!-- colnames(toolmark1.confusion)<- c("Classification", "Match", "Non-Match") -->
<!-- toolmark1.confusion$Classification<- c("Match", "Non-Match") -->
<!-- # toolmark1.confusion$Match<- c(41,2) -->
<!-- # toolmark1.confusion$`Non-Match`<- c(9,48) -->
<!-- toolmark1.confusion$Match<- c(91,2) -->
<!-- toolmark1.confusion$`Non-Match`<- c(9,98) -->

<!-- a2<- matrix(NA, ncol = 3, nrow = 2) -->
<!-- toolmark2.confusion<- data.frame(a2) -->
<!-- colnames(toolmark2.confusion)<- c("Classification", "Match", "Non-Match") -->
<!-- toolmark2.confusion$Classification<- c("Match", "Non-Match") -->
<!-- toolmark2.confusion$Match<- c(47,0) -->
<!-- toolmark2.confusion$`Non-Match`<- c(3,50) -->

<!-- chumbley.confusion<-knitr::kable(toolmark1.confusion, digits=0, format = "latex" ,booktabs=TRUE) #%>% kableExtra::kable_styling(latex_options = c("hold")) -->
<!-- hadler.confusion<-knitr::kable(toolmark2.confusion, digits=0, format = "latex" ,booktabs=TRUE) #%>% kableExtra::kable_styling(latex_options = c("hold")) -->

<!-- cat(c("\\begin{table}[!htb] -->
<!--     \\begin{minipage}{.5\\linewidth} -->
<!--       \\caption{Chumbley et al. 2010} -->
<!--       \\centering", -->
<!--         chumbley.confusion, -->
<!--     "\\end{minipage}% -->
<!--     \\begin{minipage}{.5\\linewidth} -->
<!--       \\centering -->
<!--         \\caption{Hadler et. al. (2017)}", -->
<!--        hadler.confusion, -->
<!--     "\\end{minipage}  -->
<!-- \\end{table}" -->
<!-- )) -->

@@

<!-- \citet{chumbley} found that the algorithm gives false-positives in slightly over 2 cases (or 2%) of the time and about 9 false-negatives (or 9%) for every 100 comparisons. \citet{hadler} on the other hand, in the improved version of the algorithm first described by \citet{chumbley} found false-positives for 0 cases and 3 false-negatives (or 6%) for 50 comparisons of knwon matches. -->
<!-- ```{r} -->
<!-- a<- matrix(NA, ncol = 3, nrow = 2) -->
<!-- toolmark.falsepositive<- data.frame(a) -->
<!-- colnames(toolmark.falsepositive)<- c("Classification", "Match", "Non-Match") -->
<!-- toolmark.falsepositive$Classification<- c("Match", "Non-Match") -->
<!-- toolmark.falsepositive$Match<- c(47,0) -->
<!-- toolmark.falsepositive$`Non-Match`<- c(3,50) -->

<!-- knitr::kable(toolmark.falsepositive, digits=0, format = "latex" ,booktabs=TRUE) %>% kableExtra::kable_styling(latex_options = c("hold")) -->
<!-- ``` -->

<!-- ## Potential limitations of the Chumbley Score adaptation to bullets -->
<!-- Bullet striations are different from toolmarks. Bullet markings are typically produced by the barrel on the bullets, while toolmarks can be made by tool tips on different surfaces. Tools like screw driver tips as used by \citet{chumbley}, produce longer and pronounced markings. Bullets on the other hand are much smaller in length, width, and curved in the cross-sectional topography. This means the markings made by barrels as opposed to markings made by tools, may have a problem in distinctiveness. The majority of Bullet profiles and signatures of the Hamby study \citep{hamby} used in this paper are extracted by procedures mentioned by  \citet{aoas}. The Ruger pistol barrels used in the  Hamby study have a 9mm caliber. This means that bullets used in these pistols have land markings of almost $(1/6)^{th}$ the size of screwdriver toolmarks used by \citet{chumbley}.  Also, striations on bullets are made on their curved surfaces, whereas the algorithm developed by \citet{chumbley} and \citet{hadler} has only been tested for flatter and wider surfaces which have negligible curvature. Therefore, using methods proposed for toolmarks may need adaptation in order to give tangible results for bullets. Moreover, in order to get flat bullet signatures and remove the curvatures some kind of smoothing needs to be applied as a pre-step wihch needs further investigation as to whether the level of smoothing does effect the working of the algorithm on Bullets. -->

<!-- The Chumbley method and algorithm works on the premise of comparison windows which are supposed to represent small segments of pixels of equal sizes in the two markings that are to be compared in a predetermined fashion. They term them as windows of optimization and validation. The optimization window is selected in the first step called the optimization step where areas/windows of best agreement between the the two markings are found. The window of optimization for bullet striations are shorter as bullet signatures are smaller compared to toolmarks . The idea is to keep the number of windows of optimization sufficiently large. This means we have shorter trace segments that lets us compare smaller segments of one signature to another than those for toolmarks. Here, trace segments are the partitions of marking with the length of the partition equal to the size of window of optimization. This introduces a problem as, if we go too small in the window of optimization, the unique features of the trace segments are lost and seem similar, while too large sizes vastly reduces the weight of small features that would otherwise uniquely classify a signature and hence identify the region of agreement. The problem of distinctiveness in the structure of bullet striations may also lead to potential failure of the algorithm to identify correct windows of agreements or not be able to identify these windows at all. This may lead to incorrect classification of bullets as matches or non-matches. -->

<!-- Thus, the comaparison windows have a direct influence on the false positives and false negatives, making identification of the optimum sizes of these windows very important. This raises important questions about how and what parameter settings need to be chosen for bullet land comparison. Something similar is done in \citet{afte-chumbley} for toolmark comparisons of slip-joint pliers where optimum window sizes are determined. There is a need to figure out the best parameter settings which minimizes the errors for unsmoothed markings or profiles and pre-processed signatures. An analysis of these error rates and comparison with other methods will help us understand the adaptibility of the chumbley score to bullets.   -->

@@

Initially, the default settings suggested in \citet{hadler} were used: $w_o = 120$ pixels or about 190 $\mu m$ (ten percent of the average length of profiles) and coarseness $c = 0.25$. \autoref{fig:fails} shows the percentage of failed tests among the 86k land-to-land comparisons of the NIST data for different values of the validation window size $w_v$. For same-source lands up to 12.5 percent of the tests fail.

@@

@@

>>>>

## Motivation
Same source analyses are a major part of an Forensic Toolmark Examiner's job. In current practice examiners  make these comparisons by visual inspection under a comparison microscope and come to one of the following four conclusions: identification, inconclusive, elimination or unsuitable for examination~\citep{afte-toolmarks1998}. These conclusions are made on the basis of "unique surface contours" of the two toolmarks being in "sufficient agreement" \citep{afte-toolmarks1998}. AFTE describes the term "sufficient agreement" as the possibility of another tool producing the markings under comparison, as practically impossible \citep{afte-toolmarks1998}. Potential subject bias in the assessment as well as the lack of specified error rates are the main points of criticisms  first raised by the National Research Council in 2009 \citep{NAS:2009} and later emphasized further by the President's Council of Advisors on Science and Technology \citep{pcast2016}.

@@

Same-shift pairs of length $w_v$ are defined in \citet{hadler} as all 
pairs $(s_i^{(1)}, s_i^{(2)})$ for all integer values $i$ for which both $s_i^{(1)}$ and  $s_i^{(2)}$ are defined. Similarly, different-shift pairs are defined as $(s_i^{(1)}, s_{-i}^{(2)})$ for all $i$ where both $s_i^{(1)}$ and  $s_{-i}^{(2)}$ are defined (see \autoref{sketch-same-diff}).

\begin{figure}[hbtp]
\centering
\includegraphics[width=.7\textwidth]{images/sketch-same.png}

\includegraphics[width=.7\textwidth]{images/sketch-diff.png}
\caption{\label{sketch-same-diff}Sketch of same-shift pairings  (top) and different-shift pairings (bottom). }
\end{figure}

@@

@@

<!-- In a previous study conducted by \citet{aoas} a machine learning based algorithm was developed for same source matching of bullets and error rates were discussed using the database from the Hamby Study \citep{hamby}. The method proposed by \citet{chumbley} (and later improved by \citet{hadler}) also provides a means to determine error rates and claims to reduce subject bias. Therefore, giving us a strong motivation to explore the adaptibility of the Chumbley score methodology to bullets. In this paper, we take this toolmark based methodology and try to apply and adapt it to bullets. After this, we consequently discuss about the efforts in doing so, along with the associated error rates. The data used in this paper also belongs to the Hamby Study \citep{hamby}. This gives us a common platform for comparing the performance of the chumbley method on bullets with the already existing method proposed by \citet{aoas} for bullets. As a brief overview, \citet{chumbley}, in their paper, compare two toolmarks with the intention of determining if it comes from the same tool.  They use an empirical based setup to validate their proposed algorithm and quantitative method which calculates a U-statisic for the purpose of classification of toolmarks as matching or non-matching. The data for their study was obtained from 50 sequentially manufactured screwdriver tips, and preselected comparison window sizes were given as inputs to the algorithm. The algorithm then compares the two toolmarks and comes up with a Mann-Whitney U-statistic and an associated p-value to designate them as matches or non-matches. The performance for every 100 comparisons, of the algorithm proposed by \citet{chumbley} and the improvement proposed by \citet{hadler} are listed in the table below. -->

@@

@@

Initially, the default settings suggested in \citet{hadler} were used: $w_o = 120$ pixels or about 190 $\mu m$ (ten percent of the average length of profiles) and coarseness $c = 0.25$. \autoref{fig:fails} shows the percentage of failed tests among the 85,491 land-to-land comparisons of the NIST data for different values of the validation window size $w_v$. For same-source lands up to 12.5 percent of the tests fail.

@@

Looking closer, we see that the problem lies in the way the different-shift pairs are picked: the number of possible different-shift pairs depends on the location of  the  optimal locations $t_1^o$ and $t_2^o$. Whenever are close to the boundaries, the number of possible pairings decreases and reaches zero, if $t_i^{(o)} < w_v$ or  $t_i^{(o)} > T_i-w_o- w_v$. For uniformly distributed locations $t_i^{(o)}$, this results in an expected rate of failure of $2 w_v / (T_i-w_o)$, or about 5.6\% for an average length of $T_i$ of 1200 pixels, $w_o = 120$ pixels and $w_v = 30$ pixels.

@@

\begin{figure}[hbtp]
\centering
\includegraphics[width=.7\textwidth]{images/sketch-diff-2.png}
\caption{\label{sketch-diff-2}Sketch of adjusted different-shift pairings. }
\end{figure}

### 2018-04-18 doc/paper.tex
expected failure rate

% !TeX program = pdfLaTeX
\documentclass[12pt]{article}
\usepackage{amsmath}
\usepackage{amsthm}
\usepackage{graphicx,psfrag,epsf}
\usepackage{enumerate}
\usepackage{natbib}
\usepackage{booktabs}
\usepackage{longtable}
\usepackage{array}
\usepackage{multirow}
\usepackage[table]{xcolor}
\usepackage{wrapfig}
\usepackage{float}
\usepackage{colortbl}
\usepackage{hyperref}
\usepackage{pdflscape}
\usepackage{tabu}
\usepackage{threeparttable}

@@

Same source analyses are a major part of an Forensic Toolmark Examiner's
job. In current practice examiners make these comparisons by visual
inspection under a comparison microscope and come to one of the
following four conclusions: identification, inconclusive, elimination or
unsuitable for examination\textasciitilde{}\citep{afte-toolmarks1998}.
These conclusions are made on the basis of ``unique surface contours''
of the two toolmarks being in ``sufficient agreement''
\citep{afte-toolmarks1998}. AFTE describes the term ``sufficient
agreement'' as the possibility of another tool producing the markings
under comparison, as practically impossible \citep{afte-toolmarks1998}.
This subjectivity in the assessment as well as the lack of error rates
are the main points of criticisms first raised by the National Research
Council in 2009 \citep{NAS:2009} and later emphasized further by the
President's Council of Advisors on Science and Technology
\citep{pcast2016}.

@@

Same-shift pairs of length \(w_v\) are defined in \citet{hadler} as all
pairs \((s_i^{(1)}, s_i^{(2)})\) for all integer values \(i\) for which
both \(s_i^{(1)}\) and \(s_i^{(2)}\) are defined. Similarly,
different-shift pairs are defined as \((s_i^{(1)}, s_{-i}^{(2)})\) for
all \(i\) where both \(s_i^{(1)}\) and \(s_{-i}^{(2)}\) are defined.

@@

For both samples the correlations between the pairs of markings are then
calculated. The intuition here is that for two markings from the same
source the correlation for the in-phase sample should be high, while the
correlations of the out-of-phase sample provide a measure for the
base-level correlation for non-matching marks of a given length \(w_v\).
The Chumbley score is then computed as a Mann Whitney U statistic to
compare between in-phase sample and out-of-phase sample. In the original
method proposed in \citet{chumbley} both in-phase and out-of-phase
sample are extracted randomly, whereas \citet{hadler} proposed
deterministic rules for both samples to make the resulting score
deterministic while simultaneously avoiding overlaps within selected
marks to ensure independence.

@@

Initially, the default settings suggested in \citet{hadler} were used:
\(w_o = 120\) pixels or about 190 \(\mu m\) (ten percent of the average
length of profiles) and coarseness \(c = 0.25\). \autoref{fig:fails}
shows the percentage of failed tests among the 86k land-to-land
comparisons of the NIST data for different values of the validation
window size \(w_v\). For same-source lands up to 12.5 percent of the
tests fail.

@@

>>>>

% !TeX program = pdfLaTeX
\documentclass[12pt]{article}
\usepackage{amsmath}
\usepackage{amsthm}
\usepackage{graphicx,psfrag,epsf}
\usepackage{enumerate}
\usepackage{natbib}
\usepackage{booktabs}
\usepackage{longtable}
\usepackage{array}
\usepackage{multirow}
\usepackage{subfig}
\usepackage[table]{xcolor}
\usepackage{wrapfig}
\usepackage{float}
\usepackage{colortbl}
\usepackage{hyperref}
\usepackage{pdflscape}
\usepackage{tabu}
\usepackage{threeparttable}

@@

Same source analyses are a major part of an Forensic Toolmark Examiner's
job. In current practice examiners make these comparisons by visual
inspection under a comparison microscope and come to one of the
following four conclusions: identification, inconclusive, elimination or
unsuitable for examination\textasciitilde{}\citep{afte-toolmarks1998}.
These conclusions are made on the basis of ``unique surface contours''
of the two toolmarks being in ``sufficient agreement''
\citep{afte-toolmarks1998}. AFTE describes the term ``sufficient
agreement'' as the possibility of another tool producing the markings
under comparison, as practically impossible \citep{afte-toolmarks1998}.
Potential subject bias in the assessment as well as the lack of
specified error rates are the main points of criticisms first raised by
the National Research Council in 2009 \citep{NAS:2009} and later
emphasized further by the President's Council of Advisors on Science and
Technology \citep{pcast2016}.

@@

Same-shift pairs of length \(w_v\) are defined in \citet{hadler} as all
pairs \((s_i^{(1)}, s_i^{(2)})\) for all integer values \(i\) for which
both \(s_i^{(1)}\) and \(s_i^{(2)}\) are defined. Similarly,
different-shift pairs are defined as \((s_i^{(1)}, s_{-i}^{(2)})\) for
all \(i\) where both \(s_i^{(1)}\) and \(s_{-i}^{(2)}\) are defined (see
\autoref{sketch-same-diff}).

\begin{figure}[hbtp]
\centering
\includegraphics[width=.7\textwidth]{images/sketch-same.png}

\includegraphics[width=.7\textwidth]{images/sketch-diff.png}
\caption{\label{sketch-same-diff}Sketch of same-shift pairings  (top) and different-shift pairings (bottom). }
\end{figure}

@@

For both same- and different-shift pairs correlations between the
markings are calculated. The intuition here is that for two markings
from the same source the correlation for the in-phase sample should be
high, while the correlations of the out-of-phase sample provide a
measure for the base-level correlation for non-matching marks of a given
length \(w_v\). The Chumbley score is then computed as a Mann Whitney U
statistic to compare between in-phase sample and out-of-phase sample. In
the original method proposed in \citet{chumbley} both in-phase and
out-of-phase sample are extracted randomly, whereas \citet{hadler}
proposed the above specified deterministic rules for both samples to
make the resulting score deterministic while simultaneously avoiding
overlaps within selected marks to ensure independence.

@@

Initially, the default settings suggested in \citet{hadler} were used:
\(w_o = 120\) pixels or about 190 \(\mu m\) (ten percent of the average
length of profiles) and coarseness \(c = 0.25\). \autoref{fig:fails}
shows the percentage of failed tests among the 85,491 land-to-land
comparisons of the NIST data for different values of the validation
window size \(w_v\). For same-source lands up to 12.5 percent of the
tests fail.

@@

Looking closer, we see that the problem lies in the way the
different-shift pairs are picked: the number of possible different-shift
pairs depends on the location of the optimal locations \(t_1^o\) and
\(t_2^o\). Whenever are close to the boundaries, the number of possible
pairings decreases and reaches zero, if \(t_i^{(o)} < w_v\) or
\(t_i^{(o)} > T_i-w_o- w_v\). For uniformly distributed locations
\(t_i^{(o)}\), this results in a probability of failure of
\(2 w_v / (T_i-w_o)\), or about 5.6\% for an average length of \(T_i\)
of 1200 pixels, \(w_o = 120\) pixels and \(w_v = 30\) pixels.
